Modelling circumstellar discs with 3D radiation hydrodynamics

### 2009-10-09 SPH-Torus-paper1/sphtorus_benchmark.tex
Improved figure captions and formatting.
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2048 25464f21-6099-402c-b30a-640f1d40d38b

\title{Modelling circumstellar discs with 3D radiation hydrodynamics}
\author[David M. Acreman et al]{David M. Acreman et al\\
School of Physics, University of Exeter, Stocker Road, Exeter EX4
4QL. E-mail acreman@astro.ex.ac.uk
}

@@

The density of the disc benchmark described by \cite{pascucci_2004} is
given by
\begin{equation}
\rho\left(r,z\right) = \rho_0 \left( \frac{r}{r_d} \right)^{-1} \exp
\left( - \frac{\pi}{4} \left( \frac{z}{h\left(r \right)}  \right)^2 \right)
\label{eqn:pascucci_density}
\end{equation}
where $r$ is the distance from the central star in the mid-plane and
$z$ is the height from the mid-plane. The expression $h\left( r
\right)$ is given by
\begin{equation}
h\left( r \right) = z_d \left( \frac{r}{r_d}\right)^{1.125}  
\end{equation}
and $\rho_0$, $z_d$ and $r_d$ are constant parameters.  We base our
benchmark model on the most optically thick case presented by
\cite{pascucci_2004} which has a mid-plane optical depth of $\tau=100$
at $\lambda=550~\rm{nm}$ and a mass of 0.011~$\rm{M}_{\sun}$. This
requires $r_d=500~\rm{AU}$, $z_d=0.25 r_d$ and
$\rho_{0}=8.1614\times10^{-18}~\rm{g~cm^{-3}}$, with the disc truncated at
an outer radius of 1000~AU and an inner radius of 1~AU.

@@

To illustrate the effects of the different splitting methods two
example grids are shown in Fig.~\ref{fig:example_meshes}. A grid
constructed from $10^7$ particles, using a mass per cell limit of
$5\times10^{26}~\rm{g}$, is plotted in
Fig.~\ref{fig:mass_condition_mesh}. This plot is a slice perpendicular
to the disc mid-plane and shows increased resolution towards the disc
centre and mid-plane where the density is higher.  A grid with
$f_{\rm{split}}=0.1$, also using $10^7$ particles, is plotted in
Fig.~\ref{fig:density_condition_mesh}. The resolution increases
towards the edge of the disc, where there are large density gradients,
and also increases towards the centre of the disc. 
\begin{figure*}
  \subfigure[AMR grid with a mass per cell condition of
    $5\times10^{26}~\rm{g}$.]
    {\includegraphics[scale=0.22]{figures/set1_mesh}
  \label{fig:mass_condition_mesh}}
  \subfigure[Grid with a density condition of
    $f_{\rm{split}}=0.1$.]
    {\includegraphics[scale=0.22]{figures/set2_mesh}
      \label{fig:density_condition_mesh}}
  \caption{Example AMR grids for a mass per cell splitting condition (left)
    and a density contrast splitting condition (right). In both cases
    $10^7$ particles were used. The axis units are $10^{16}$~cm.}
  \label{fig:example_meshes}
\end{figure*}

For each splitting condition a number of AMR grids were generated
using different values of the mass per cell limit or
$f_{\rm{split}}$. This was repeated for discs represented by $10^5$,
$10^6$ and $10^7$ SPH particles. Figure~\ref{fig:grid_tests} plots the
number of octals generated as a function of mass per cell or
$f_{\rm{split}}$ (solid line), and the percentage error in the disc
mass (dashed line). The percentage error in disc mass is calculated by
comparing the total mass on the AMR grid with the known mass of
$0.011~\rm{M}_{\sun}$ based on the analytical form of the density
distribution, and a positive value indicates that mass on the grid is
too large.
\begin{figure*}
  \subfigure[Mass per cell condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_set1}}
  \subfigure[Mass per cell condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_set1}}
  \subfigure[Mass per cell condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_set1}}
  \subfigure[Density contrast condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_set2}}
  \subfigure[Density contrast condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_set2}}
  \subfigure[Density contrast condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_set2}}
  \caption{Number of octals (solid line) and percentage mass error
    (dashed line) for Cartesian AMR grids generated using different grid
    splitting methods and different numbers of particles. }
  \label{fig:grid_tests}
\end{figure*}

@@

The density contrast condition is effective in adding extra resolution
in regions of high density contrast which are likely to be important
in radiative transfer calculations. However the density split
condition has a less robust total mass representation than the mass
per cell limit. By using a combination of these two conditions it is
possible to add extra resolution in high contrast regions and still
maintain an accurate total mass.

@@

For some geometries it may be appropriate to use a cylindrical polar
co-ordinate system. A mass per cell splitting condition can still be
used but it needs to be modified to take into account the increase in 
cell volume as cylindrical polar radius increases, for fixed values of
$dr$ and $dz$.  
%
Consider a Cartesian grid with cubic volume elements and let
\begin{equation}
dl_{\rm{cart}} = dx = dy = dz 
\end{equation}
so that
\begin{equation}
dl_{\rm{cart}} = \left( \frac{dM_{\rm{cart}}}{\rho_{\rm{cell}}} \right)
^{1/3}
\label{eqn:cart_cell}
\end{equation}
where $dM_{\rm{cart}}$ is the mass of a cell in the Cartesian grid and
$\rho_{\rm{cell}}$ is the density of the cell. 
%
In a cylindrical polar grid the mass of a cell is given by
\begin{equation}
dM_{\rm{cyl}} = \rho_{\rm{cell}} \frac{2 \pi r dr dz}{n_{\rm{az}}}
\end{equation}
if there are $n_{\rm{az}}$ uniform azimuthal cells. If the 
resolution in $r$ and $z$ is the same then we can define 
\begin{equation}
dl_{\rm{cyl}} = dr = dz 
\end{equation}
which is related to the cell mass and density by
\begin{equation}
dl_{\rm{cyl}} = \left( \frac{n_{\rm{az}} dM_{\rm{cyl}}}{2 \pi r
  \rho_{\rm{cell}}} \right)  ^{1/2}
\label{eqn:cyl_cell}
\end{equation}
If we require the resolution in cylindrical polar $r$ and $z$
to match the resolution in Cartesian $x$, $y$ and $z$ then $dl_{\rm{cyl}} =
dl_{\rm{cart}}$. From eqn~\ref{eqn:cart_cell} and \ref{eqn:cyl_cell} 
\begin{equation}
dM_{\rm{cyl}} = \frac{2 \pi r}{n_{\rm{az}}} \rho^{1/3} dM_{\rm{cart}}^{2/3}
\end{equation}
which describes the required conversion from a mass per cell limit
suitable for a Cartesian grid ($dM_{\rm{cart}}$) to an equivalent mass
per cell limit suitable for a cylindrical polar grid ($dM_{\rm{cyl}}$).

The grid generation tests were run using a cylindrical polar geometry
with the modified mass per cell limit described above and an
unmodified density contrast condition. The results are plotted in
Fig.~\ref{fig:cp_grid_tests}.
\begin{figure*}
  \subfigure[Mass per cell condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_cyl_set1}}
  \subfigure[Mass per cell condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_cyl_set1}}
  \subfigure[Mass per cell condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_cyl_set1}}
  \subfigure[Density contrast condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_cyl_set2}}
  \subfigure[Density contrast condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_cyl_set2}}
  \subfigure[Density contrast condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_cyl_set2}}
  \caption{Number of octals (solid line) and percentage mass error
    (dashed line) for cylindrical polar AMR grids generated using different grid
    splitting methods and different numbers of particles. }
  \label{fig:cp_grid_tests}
\end{figure*}
Using a cylindrical polar geometry results in a significantly reduced
number of octals in the AMR grid but with larger mass errors. Provided
the mass per cell limit is no more than $10^{27}\rm{g}$ then the total
mass is well represented with significantly fewer cells than in a
Cartesian grid. The density contrast condition alone does not
represent the total mass to within 3\% accuracy even with
$f_{\rm{split}}=0.1$. Although this condition successfully adds
enhanced resolution in regions of higher density contrast it needs to
be combined with a mass per cell limit to ensure a robust total mass
representation.

@@

The disc was initially modelled using $10^5$ SPH particles with an AMR
grid constructed from a mass per cell limit of $5\times10^{26}$g.
This number of SPH particles would normally be considered as
sufficient to adequately represent a disc of this mass in a purely
hydrodynamical calculation. The mass per cell limit was chosen in
preference to the density contrast limit to give the highest
resolution near the central region of the disc, which we expect to be
most important for determining the temperature of the disc at the
inner edge and the SED.

@@

\begin{figure}
  \includegraphics[scale=0.3]{figures/nomod_tem_diff}
  \caption{Temperature difference between SPH benchmark disc and high
    accuracy 2D result.}
  \label{fig:nomod_tem_diff}
\end{figure}

SEDs were calculated using the {\sc torus} code for viewing
angles of 12.5 and 77.5 degrees (where an inclination angle of zero
degrees indicates that the disc is viewed face-on) and the results are
plotted in Fig.~\ref{fig:seds} (dashed line). Also plotted in
Fig.~\ref{fig:seds} are the benchmark results of \cite{pascucci_2004}
(solid line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/sed_013}
  \includegraphics[scale=0.3]{figures/sed_077}
  \caption{SEDs from the benchmark disc with $10^5 $ particles. SEDs
    are shown for inclination angles of 12.5 degrees (left) and 77.5
    degrees (right). The benchmark SED is plotted as a solid line and
    the {\sc torus} SED is plotted as a dashed line.}
  \label{fig:seds}
\end{figure*}
Both SEDs show significant departures from the benchmark results;
there is excess emission between 10-100 $\mu\rm{m}$ and at a viewing
angle of 77.5 degrees there is also a significant reduction in flux
below 1~$\mu\rm{m}$. Both these discrepancies are attributable to a lack of resolution within a few au of the central object. The density distribution in the central region of
this disc is plotted in Fig.~\ref{fig:basic_dens} which shows that the
central 1\,au gap is not represented. There are two effects in
operation; firstly the grid resolution in this region is larger than
1\,au  and secondly the smoothing lengths of the particles are larger
than 1\,au. In order to correctly represent the central gap there needs
to be higher grid resolution around the central source and more SPH
particles are required so that the smoothing lengths are smaller.
\begin{figure}
\includegraphics[scale=0.2]{figures/basic_dens}
\caption{Density in central region of disc with $10^5$ particles. Axis
units are $10^{13}$cm.}
\label{fig:basic_dens}
\end{figure}

\subsection{Grid resolution}
Decreasing the mass per cell limit to achieve the required resolution
around the central source would result in too many cells being
generated in other regions of the disc. In order to achieve a grid
resolution of less than 1\,au around the source it was necessary to
specify additional criteria for refining the grid. Three cubes of
increased resolution were added to the AMR grid with sizes
$1.5\times10^{15}$~cm, $3\times10^{14}$~cm and
$6\times10^{13}$~cm. Within each cube it was required that the cell
size be no larger than 0.1 of the cube dimension. In order to confirm
that the grid modifications provide sufficient resolution to generate
an accurate SED three calculations were performed with the central
part of the disc forced to the correct density values from the
analytical density description.  A spherical region with radius
$1.0\times10^{15}$~cm, $5.0\times10^{14}$~cm or $2.5\times10^{14}$~cm
was over written with the benchmark disc density after the initial
grid generation was performed. The resulting SEDs are plotted in
Fig.~\ref{fig:sed_forced}. The solid line shows the benchmark result
and the other lines show SEDs for forcing regions of radius
$1.0\times10^{15}$~cm (long dashed line), $5.0\times10^{14}$~cm (short
dashed line) and $2.5\times10^{14}$~cm (dotted line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/seds_forced_13.pdf}
  \includegraphics[scale=0.3]{figures/seds_forced_77.pdf}
  \caption{SEDs from SPH benchmark disc for inclination angles of 12.5
    degrees and 77.5 degrees. The benchmark SED is plotted as a solid
    line. The {\sc torus} density distribution is forced to the correct
    value with a radius of $1.0\times10^{15}$~cm (long dashed line), $5.0\times10^{14}$~cm
(short dashed line) and $2.5\times10^{14}$~cm (dotted line).} 
  \label{fig:sed_forced}
\end{figure*}
With the correct density forced out to a radius of
$1.0\times10^{15}$~cm the SED is close to the benchmark result for
both viewing angles, however if the forced region is reduced in size
there is a significant discrepancy relative to the benchmark SED. This
confirms that the grid now has enough resolution to generate an
accurate SED but also indicates that using $10^5$ particles does not
give a sufficiently accurate density distribution within the central
$1.0\times10^{15}$~cm of the disc.

\subsection{Number of particles}
Further calculations were performed using different numbers of SPH
particles and the enhanced grid resolution described above in order to
determine the impact of the number of SPH particles on the accuracy of
the radiative transfer calculation. The density distribution in the
central region of the disc is plotted in
Fig.~\ref{fig:n_part_central_dens} in order to show how the
representation improves as the number of particles is increased.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.22]{figures/dens_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.22]{figures/dens_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.22]{figures/dens_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.22]{figures/dens_1e8}}
  \caption{Central density on AMR grid derived from different numbers
    of SPH particles. The axis units are $10^{13}$~cm.} 
  \label{fig:n_part_central_dens}
\end{figure*}
The effective resolution of the AMR mesh is dependent on two factors; the probability of having sufficient particles to adequately sample the small volume of the disc inner-edge, and the particle smoothing length (which determines the effective spatial resolution of the SPH density representation).
With $10^5$ and $10^6$ particles a central gap is present but is too
large. With $10^7$ particles there is more material close in to the
source but the gap itself is not represented (because the smoothing length of the particles near the disc inner-edge is too large) and the source is
obscured. Only with $10^8$ particles is a gap of approximately the
correct size present.  The representation of the central 1\,au gap is
significantly affected by the number of particles used to represent
the disc, with a large number of particles required to achieve an
accurate density distribution in this spatially small region.

Figure~\ref{fig:n_part_tem_diff} shows fractional errors in the
temperature distribution for different numbers of
particles. 
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e8}}
  \caption{Fractional temperature error for discs represented by
    different numbers of SPH particles. Axis units are au.} 
  \label{fig:n_part_tem_diff}
\end{figure*}
When using $10^5$ particles to represent the disc there is a too much
heating in the disc mid plane and insufficient heating further out of
the mid plane. The effect is reduced as the number of particles is
increased and reaches a very low level with $10^8$ particles, at which stage the
temperature distribution agrees with the benchmark to within $\sim$10\%.

To allow a more quantitative comparison the mid-plane temperature and
the benchmark mid-plane temperature are plotted in
Fig.~\ref{fig:n_part_tem_diff_midpane} for all particles within 0.01
scale heights of the mid-plane. The fractional temperature error for
each particle is also plotted where a positive error indicates a
temperature hotter than the benchmark value. 
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/mp_temp_1}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/mp_temp_2}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/mp_temp_3}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/mp_temp_4}}
  \caption{Fractional temperature error in the mid-plane for discs represented by
    different numbers of SPH particles.} 
  \label{fig:n_part_tem_diff_midpane}
\end{figure*}

>>>>

\title{Modelling circumstellar discs with 3D radiation hydrodynamics}
\author[David M. Acreman, Tim J. Harries and David A. Rundle]{David M. Acreman, Tim J. Harries and David A. Rundle\\
School of Physics, University of Exeter, Stocker Road, Exeter EX4
4QL. E-mail acreman@astro.ex.ac.uk
}

@@

The density of the disc benchmark described by \cite{pascucci_2004} is
given by
\begin{equation}
\rho\left(r,z\right) = \rho_0 \left( \frac{r}{r_d} \right)^{-1} \exp
\left( - \frac{\pi}{4} \left( \frac{z}{h\left(r \right)}  \right)^2 \right)
\label{eqn:pascucci_density}
\end{equation}
where $r$ is the distance from the central star in the mid-plane and
$z$ is the height from the mid-plane. The expression $h\left( r
\right)$ is given by
\begin{equation}
h\left( r \right) = z_d \left( \frac{r}{r_d}\right)^{1.125}  
\end{equation}
and $\rho_0$, $z_d$ and $r_d$ are constant parameters.  We base our
benchmark model on the most optically thick case presented by
\cite{pascucci_2004} which has a mid-plane optical depth of $\tau=100$
at $\lambda=550\,\rm{nm}$ and a mass of 0.011\,$\rm{M}_{\sun}$. This
requires $r_d=500\,\rm{au}$, $z_d=0.25 r_d$ and
$\rho_{0}=8.1614\times10^{-18}\,\rm{g\,cm^{-3}}$, with the disc truncated at
an outer radius of 1000\,au and an inner radius of 1\,au.

@@

To illustrate the effects of the different splitting methods two
example grids are shown in Fig.~\ref{fig:example_meshes}. A grid
constructed from $10^7$ particles, using a mass per cell limit of
$5\times10^{26}\,\rm{g}$, is plotted in
Fig.~\ref{fig:mass_condition_mesh}. This plot is a slice perpendicular
to the disc mid-plane and shows increased resolution towards the disc
centre and mid-plane where the density is higher.  A grid with
$f_{\rm{split}}=0.1$, also using $10^7$ particles, is plotted in
Fig.~\ref{fig:density_condition_mesh}. The resolution increases
towards the edge of the disc, where there are large density gradients,
and also increases towards the centre of the disc. 
\begin{figure*}
  \subfigure[AMR grid with a mass per cell condition of
    $5\times10^{26}\,\rm{g}$.]
    {\includegraphics[scale=0.22]{figures/set1_mesh}
  \label{fig:mass_condition_mesh}}
  \subfigure[Grid with a density condition of
    $f_{\rm{split}}=0.1$.]
    {\includegraphics[scale=0.22]{figures/set2_mesh}
      \label{fig:density_condition_mesh}}
  \caption{Example AMR grids for a mass per cell splitting condition
    of  $5\times10^{26}\,\rm{g}$ (left)
    and a density contrast splitting condition of $f_{\rm{split}}=0.1$
    (right) contructed
    from $10^7$ SPH particles. The grids are
    slices perpendicular to the disc mid-plane and the axis units are $10^{16}$\,cm.}
  \label{fig:example_meshes}
\end{figure*}

For each splitting condition a number of AMR grids were generated
using different values of the mass per cell limit or
$f_{\rm{split}}$. This was repeated for discs represented by $10^5$,
$10^6$ and $10^7$ SPH particles. Figure~\ref{fig:grid_tests} plots the
number of octals generated as a function of mass per cell or
$f_{\rm{split}}$ (solid line), and the percentage error in the disc
mass (dashed line). The percentage error in disc mass is calculated by
comparing the total mass on the AMR grid with the known mass of
$0.011\,\rm{M}_{\sun}$ based on the analytical form of the density
distribution, and a positive value indicates that mass on the grid is
too large.
\begin{figure*}
  \subfigure[Mass per cell condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_set1}}
  \subfigure[Mass per cell condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_set1}}
  \subfigure[Mass per cell condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_set1}}
  \subfigure[Density contrast condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_set2}}
  \subfigure[Density contrast condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_set2}}
  \subfigure[Density contrast condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_set2}}
  \caption{Number of octals (solid line) and percentage error in total
    disc mass (dashed line) for the benchmark disc represented on
    Cartesian AMR grids generated using different grid splitting
    methods (mass per cell limit or density contrast limit) and
    different numbers of particles ($10^5$, $10^6$ or $10^7$). }
  \label{fig:grid_tests}
\end{figure*}

@@

The density contrast condition is effective in adding extra resolution
in regions of high density contrast which are likely to be important
in radiative transfer calculations. However this condition has a less
robust total mass representation than the mass per cell limit. By
using a combination of these two conditions it is possible to add
extra resolution in high contrast regions and still maintain an
accurate total mass.

@@

For some geometries it may be appropriate to use a cylindrical polar
co-ordinate system. A mass per cell splitting condition can still be
used but needs to be modified to take into account that for fixed
values of $dr$ and $dz$ the cell volume increases as cylindrical polar
radius increases.
%
Consider a Cartesian grid with cubic volume elements and let
\begin{equation}
dl_{\rm{cart}} = dx = dy = dz 
\end{equation}
so that
\begin{equation}
dl_{\rm{cart}} = \left( \frac{dM_{\rm{cart}}}{\rho_{\rm{cell}}} \right)
^{1/3}
\label{eqn:cart_cell}
\end{equation}
where $dM_{\rm{cart}}$ is the mass of a cell in the Cartesian grid and
$\rho_{\rm{cell}}$ is the density of the cell. 
%
In a cylindrical polar grid the mass of a cell is given by
\begin{equation}
dM_{\rm{cyl}} = \rho_{\rm{cell}} \frac{2 \pi r dr dz}{n_{\rm{az}}}
\end{equation}
if there are $n_{\rm{az}}$ uniform azimuthal cells. If the 
resolution in $r$ and $z$ is the same then we can define 
\begin{equation}
dl_{\rm{cyl}} = dr = dz 
\end{equation}
which is related to the cell mass and density by
\begin{equation}
dl_{\rm{cyl}} = \left( \frac{n_{\rm{az}} dM_{\rm{cyl}}}{2 \pi r
  \rho_{\rm{cell}}} \right)  ^{1/2}
\label{eqn:cyl_cell}
\end{equation}
If we require the resolution in cylindrical polar $r$ and $z$
to match the resolution in Cartesian $x$, $y$ and $z$ then $dl_{\rm{cyl}} =
dl_{\rm{cart}}$. From eqn~\ref{eqn:cart_cell} and \ref{eqn:cyl_cell} 
\begin{equation}
dM_{\rm{cyl}} = \frac{2 \pi r}{n_{\rm{az}}} \rho^{1/3} dM_{\rm{cart}}^{2/3}
\end{equation}
which describes the required conversion from a mass per cell limit
suitable for a Cartesian grid ($dM_{\rm{cart}}$) to an equivalent mass
per cell limit suitable for a cylindrical polar grid ($dM_{\rm{cyl}}$).

The grid generation tests were run using a cylindrical polar geometry
with the modified mass per cell limit described above and an
unmodified density contrast condition. The results are plotted in
Fig.~\ref{fig:cp_grid_tests}.
\begin{figure*}
  \subfigure[Mass per cell condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_cyl_set1}}
  \subfigure[Mass per cell condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_cyl_set1}}
  \subfigure[Mass per cell condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_cyl_set1}}
  \subfigure[Density contrast condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_cyl_set2}}
  \subfigure[Density contrast condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_cyl_set2}}
  \subfigure[Density contrast condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_cyl_set2}}
  \caption{Number of octals (solid line) and percentage error in total
    disc mass (dashed line) for the benchmark disc represented on
    cylindrical polar AMR grids generated using different grid splitting
    methods (mass per cell limit or density contrast limit) and
    different numbers of particles ($10^5$, $10^6$ or $10^7$).}
  \label{fig:cp_grid_tests}
\end{figure*}
Using a cylindrical polar geometry results in a significantly reduced
number of octals in the AMR grid but with larger mass errors. Provided
the mass per cell limit is no more than $10^{27}\rm{g}$ then the total
mass is well represented with significantly fewer cells than in a
Cartesian grid. The density contrast condition alone does not
represent the total mass to within 3\% accuracy even with
$f_{\rm{split}}=0.1$. Although this condition successfully adds
enhanced resolution in regions of higher density contrast it needs to
be combined with a mass per cell limit to ensure a robust total mass
representation.

@@

The disc was initially modelled using $10^5$ SPH particles with an AMR
grid constructed from a mass per cell limit of $5\times10^{26}$\,g.
This number of SPH particles would normally be considered as
sufficient to adequately represent a disc of this mass in a purely
hydrodynamical calculation. The mass per cell limit was chosen in
preference to the density contrast limit to give the highest
resolution near the central region of the disc, which we expect to be
most important for determining the temperature of the disc at the
inner edge and the SED.

@@

\begin{figure}
  \includegraphics[scale=0.3]{figures/nomod_tem_diff}
  \caption{Fractional errors in the temperature distribution of a
    $10^5$ particle disc. The radiative transfer grid is a Cartesian
    grid generated using a mass per cell limit of
    $5\times10^{26}$\,g. Axis units are au.}
  \label{fig:nomod_tem_diff}
\end{figure}

SEDs were calculated using the {\sc torus} code for viewing
angles of 12.5 and 77.5 degrees (where an inclination angle of zero
degrees indicates that the disc is viewed face-on) and the results are
plotted in Fig.~\ref{fig:seds} (dashed line). Also plotted in
Fig.~\ref{fig:seds} are the benchmark results of \cite{pascucci_2004}
(solid line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/sed_013}
  \includegraphics[scale=0.3]{figures/sed_077}
  \caption{SEDs from the benchmark disc represented by $10^5 $ particles and a
    Cartesian radiative transfer grid generated using a mass per cell
    limit of $5\times10^{26}$\,g. SEDs are shown for inclination
    angles of 12.5 degrees (left) and 77.5 degrees (right). The
    benchmark SED is plotted as a solid line and the {\sc torus} SED
    is plotted as a dashed line. The corresponding fractional
    temperature errors are plotted in Fig.\ref{fig:nomod_tem_diff}.}
  \label{fig:seds}
\end{figure*}
Both SEDs show significant departures from the benchmark results;
there is excess emission between 10-100 $\mu\rm{m}$ and at a viewing
angle of 77.5 degrees there is also a significant reduction in flux
below 1\,$\mu\rm{m}$. Both these discrepancies are attributable to a lack of resolution within a few au of the central object. The density distribution in the central region of
this disc is plotted in Fig.~\ref{fig:basic_dens} which shows that the
central 1\,au gap is not represented. There are two effects in
operation; firstly the grid resolution in this region is larger than
1\,au  and secondly the smoothing lengths of the particles are larger
than 1\,au. In order to correctly represent the central gap there needs
to be higher grid resolution around the central source and more SPH
particles are required so that the smoothing lengths are smaller.
\begin{figure}
\includegraphics[scale=0.2]{figures/basic_dens}
\caption{Density in the central region of the benchmark disc
  represented by $10^5$ particles and a Cartesian radiative transfer
  grid generated using a mass per cell limit of
  $5\times10^{26}$\,g. Axis units are $10^{13}$cm. The central gap of
  1\,au ($1.5\time10^{13}$\,cm) is not represented and the effects on
  the temperature distribution and emergent SEDs are shown in
  Fig.~\ref{fig:nomod_tem_diff} and Fig.~\ref{fig:seds}
  respectively.}
\label{fig:basic_dens}
\end{figure}

\subsection{Grid resolution}
\label{subsec:grid_res}

Decreasing the mass per cell limit to achieve the required resolution
around the central source would result in too many cells being
generated in other regions of the disc. In order to achieve a grid
resolution of less than 1\,au around the source it was necessary to
specify additional criteria for refining the grid. Three cubes of
increased resolution were added to the AMR grid with sizes
$1.5\times10^{15}$\,cm, $3\times10^{14}$\,cm and
$6\times10^{13}$\,cm. Within each cube it was required that the cell
size be no larger than 0.1 of the cube dimension. In order to confirm
that the grid modifications provide sufficient resolution to generate
an accurate SED three calculations were performed with the central
part of the disc forced to the correct density values from the
analytical density description.  A spherical region with radius
$1.0\times10^{15}$\,cm, $5.0\times10^{14}$\,cm or $2.5\times10^{14}$\,cm
was over written with the benchmark disc density after the initial
grid generation was performed. The resulting SEDs are plotted in
Fig.~\ref{fig:sed_forced}. The solid line shows the benchmark result
and the other lines show SEDs for forcing regions of radius
$1.0\times10^{15}$\,cm (long dashed line), $5.0\times10^{14}$\,cm (short
dashed line) and $2.5\times10^{14}$\,cm (dotted line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/seds_forced_13.pdf}
  \includegraphics[scale=0.3]{figures/seds_forced_77.pdf}
  \caption{ SEDs from the benchmark disc represented by $10^5 $ particles and
    a Cartesian radiative transfer grid generated using a mass per
    cell limit of $5\times10^{26}$\,g. Extra resolution has been added
    around the central source and the {\sc torus} density distribution
    is forced to the correct value within a radius of
    $1.0\times10^{15}$\,cm (long dashed line), $5.0\times10^{14}$\,cm
    (short dashed line) and $2.5\times10^{14}$\,cm (dotted line). SEDs
    are shown for inclination angles of 12.5 degrees (left) and 77.5
    degrees (right) with the benchmark result plotted as a solid
    line.}
  \label{fig:sed_forced}
\end{figure*}
With the correct density forced out to a radius of
$1.0\times10^{15}$\,cm the SED is close to the benchmark result for
both viewing angles, however if the forced region is reduced in size
there is a significant discrepancy relative to the benchmark SED. This
confirms that the grid now has enough resolution to generate an
accurate SED but also indicates that using $10^5$ particles does not
give a sufficiently accurate density distribution within the central
$1.0\times10^{15}$\,cm of the disc.

\subsection{Number of particles}
Further calculations were performed using different numbers of SPH
particles and the enhanced grid resolution described above in order to
determine the impact of the number of SPH particles on the accuracy of
the radiative transfer calculation. The density distribution in the
central region of the disc is plotted in
Fig.~\ref{fig:n_part_central_dens} in order to show how the
representation improves as the number of particles is increased.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.22]{figures/dens_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.22]{figures/dens_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.22]{figures/dens_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.22]{figures/dens_1e8}}
  \caption{Density in the central region of the disc as represented on AMR grids derived from different numbers
    of SPH particles. The AMR grid is generated using a mass per cell
    limit of $5\times10^{26}$\,g with extra resolution at the origin
    as described in Section~\ref{subsec:grid_res}. The axis units are $10^{13}$\,cm.}
  \label{fig:n_part_central_dens}
\end{figure*}
The effective resolution of the AMR mesh is dependent on two factors; the probability of having sufficient particles to adequately sample the small volume of the disc inner-edge, and the particle smoothing length (which determines the effective spatial resolution of the SPH density representation).
With $10^5$ and $10^6$ particles a central gap is present but is too
large. With $10^7$ particles there is more material close in to the
source but the gap itself is not represented (because the smoothing length of the particles near the disc inner-edge is too large) and the source is
obscured. Only with $10^8$ particles is a gap of approximately the
correct size present.  The representation of the central 1\,au gap is
significantly affected by the number of particles used to represent
the disc, with a large number of particles required to achieve an
accurate density distribution in this spatially small region.

Figure~\ref{fig:n_part_tem_diff} shows fractional errors in the
temperature distribution for different numbers of
particles. 
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e8}}
  \caption{Fractional errors in the temperature distribution of discs
    represented by $10^5$, $10^6$, $10^7$ and $10^8$ particles. The
    AMR grid is generated using a mass per cell limit of
    $5\times10^{26}$\,g with extra resolution at the origin as
    described in Section~\ref{subsec:grid_res}. The axis units are
    $10^{13}$\,cm. The corresponding density distributions in the
    central part of the disc are plotted in
    Fig.~\ref{fig:n_part_central_dens}.}
  \label{fig:n_part_tem_diff}
\end{figure*}
When using $10^5$ particles to represent the disc there is a too much
heating in the disc mid plane and insufficient heating further out of
the mid plane. The effect is reduced as the number of particles is
increased and reaches a very low level with $10^8$ particles, at which stage the
temperature distribution agrees with the benchmark to within $\sim$10\%.

To allow a more quantitative comparison the mid-plane temperature and
the benchmark mid-plane temperature are plotted in
Fig.~\ref{fig:n_part_tem_diff_midpane} for all particles within 0.01
scale heights of the mid-plane. The fractional temperature error for
each particle is also plotted where a positive error indicates a
temperature hotter than the benchmark value. 
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/mp_temp_1}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/mp_temp_2}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/mp_temp_3}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/mp_temp_4}}
  \caption{Fractional temperature error, as a function of radius, for particles within 0.01
    scale heights of the mid-plane for discs represented by $10^5$,
    $10^6$, $10^7$ and $10^8$ particles. }
  \label{fig:n_part_tem_diff_midpane}
\end{figure*}
